### 2011-12-17 paper_draft.tex
added some text, moved outline to outline.txt

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome wide association studies (GWAS) are a cheap and quick way to assess health risks by comparing Single Nucleotide Polymorphisms (SNPs) between groups of participants. Direct-To-Consumer Companies like 23andme offer their customers to sequence their SNPs alongside with an evaluation of the customer's genetic risks. However, the data 23andme and other companies generate is not accessible for other scientists, and withholds some information from their customers for various reasons. In this paper, we present an open approach to GWAS by introducing openSNP, a webpage which allows GWAS-customers to openly share their SNPs with scientists for free.  % misses survey

@@

Paragraph1:

Investigate the link between genetic variation and phenotypic change
Important because of desease stuff.

Tool used: Genome Wide Association Studies.
--> Describe in one sentence.
--> 2005 first study
--> Today 1200 GWAS performed

genetic markers usually SNPs.
SNPs definition. Allele definition.
SNP can be linked to a phenotypic trait (around 5000 linked to desease)

Paragraph2:

How to obtain data for GWAS?
--> SNPChip, used in many different species
--> Now different companies for DTC genetic testing.
--> Test results provided to customer and raw data
--> Raw data can be used for GWAS, but closed

Paragraph3:

RAW data from 1000 of individuals could be used for multiple scientific
purposes
---> Large potential dataset Are the numbers available for dataset size used
in scientic studies? 2005 93, 2008 500, 2011 23andme 100000? Plot?

Resolution of the problem: Open Science
Create repository for sharing this information. openSNP

Paragraph4:

Describe openSNP:
Open Source, non-commercial
Open Data
Community driven --> Crowd sourcing

Parses data from 23andMe deCODEme....
--> Publications crawler (Maybe mention mendeley price?)
        --PLOS and Mendeley

Advantages for scientists: See paragraph 3

Advantages for customers: 
-->Make sense of the data
-->Contribute to the public domain

@@

Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data. This is mainly due to privacy concerns of the customers. Nevertheless there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like GitHub. While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public. While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data. %WHY?

A possible solution to this can be a community-driven platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent. We designed a survey to assess interest in such a crowd sourcing platform, in which we asked how many people would be willing to share their genetic and phenotypic information with the public.

@@

\subsection*{Survey on Sharing Genetic Information}
229 participants filled out the survey. Of these 229, 180 had a self-reported chromosomal sex of XY and 56 carried a self-reported chromosomal sex of XX. The mean age of the participants is 33 (SD = 11,29) and over 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already customer of at least one DTC genetic testing company and further 30.1 \% of them plan to become one in the future. 29.7 \% don't plan to become a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).

67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would share if the company does not share the data with third parties. 6.6 \% of the participants would not share their data. There is no significant difference between sharing-habits between both chromosomal sexes (Somers-d). Those who are a customer of a DTC company or are planning to become one in the future are more likely to share their results, as compared to those who don't plan to get themselves genotyped (Somers-d).

\subsection*{Introducing openSNP}
We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license. Users are encouraged to list as many phenotypes as possible due to a simple achievement-system similar to systems found in computer games.

\section*{Discussion}
to-do list together with outlook of personalised medicine

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome wide association studies (GWAS) are a cheap and quick way to assess health risks by comparing Single Nucleotide Polymorphisms (SNPs) between groups of participants. Direct-To-Consumer Companies like 23andme offer their customers to sequence their SNPs alongside with an evaluation of the customer's genetic risks. However, the data 23andme and other companies generate is not accessible for other scientists, and withholds some information from their customers for various reasons. In this paper, we present an open approach to GWAS by introducing openSNP, a web platform which allows GWAS-customers to openly share their SNPs with scientists for free.  % misses survey

@@

@@

Although companies like 23andMe are willing to contribute to science it is not easy for individual scientists to access the data. This is mainly due to privacy concerns of the customers. Nevertheless there are individual customers who are willingly sharing their data. Most do so by uploading their data to their personal website or to open software repositories like GitHub. While this is makes it possible for scientists to access the data, it requires a lot of work to keep track of all new genotyping data that is available to the public. While projects like the SNPedia try to keep track of all the files, this still does not allow to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the Personal Genome Project, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.

A possible solution to this can be a community-driven platform that aggregates genetical and phenotypical information of people who are willing to share their data with the general public and have given their informed consent. We designed a survey to assess interest in such a crowd sourcing platform, in which we asked how many people would be willing to share their genetic and phenotypic information with the public. Additionally we built a platform which allows customers of DTC genetic testing to publicate of genetic and phenotypic information and gives researchers multiple ways to reuse the data.

@@

\subsection*{Survey on Sharing Genetic Information}
229 people, 180 with a self-reported chromosomal sex of XY, 56 with a self-reported chromosomal sex of XX, participated in the survey. The mean age of the participants is 33 (SD = 11,29) and over 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already customer of at least one DTC genetic testing company and further 30.1 \% of them plan to become one in the future. 29.7 \% don't plan to become a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Somers-d).

67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so, if the company does not share the data with third parties. 6.6 \% of the participants would not share their data. There is no significant difference between sharing-habits between both chromosomal sexes (Somers-d). Those who are a customer of a DTC company or are planing to become one in the future are more likely to share their results, compared to those who don't plan to get themselves genotyped (Somers-d).

There are significant differences, tested by Tukey's HSD test, between those people who are already genotyped and those who don't plan to get genotyped: The first group is more likely to agree to share their information because they want to help scientists (mean difference = 0.465, SE = 0.128, p = 0.001), because they think of personal benefits (mean difference = 0.448, SE = 0.183, p = 0.04) and because they are curious (mean difference = 1.159, SE = 0.193, p < 0.001).

On the other hand those hand those people who are not planning to get genotyped are more likely to not share their data, because they agree to fear discrimination (mean difference = 1.060, SE = 0.195, p < 0.001), because they agree that they feel it is a breach of their privacy (mean difference = 0.821, SE = 0.225, p = 0.001), because agree that they fear negative consequences for their family (mean difference = 0.733, SE = 0.21, p = 0.002) or because they agree that they fear personalized advertising (mean difference = 0.848, SE = 0.208, p < 0.001).

Similarly those people who would share data with their DTC provider are more likely to agree on sharing the data, because they want to help scientists (mean difference = 1.57, SE = 0.199, p < 0.001), because they think of personal benefits (mean difference = 0.951, SE = 0.308, p = 0.006), and because they are curious (mean difference = 1.99, SE = 0.321, p < 0.001).

Those participants who are not planning to get genotyped are more likely to agree to not share their data, because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).

\subsection*{openSNP}
We've created openSNP, a website which allows users to upload their genotypings from the companies 23andme, deCODEme and Family Tree under the Creative Commons Zero-license, which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - allows a complete reuse of the data without any constraints. Users are encouraged to list as many phenotypes as possible through to a simple achievement-system which rewards users that upload their data and enter phenotypic information with small badges for their efforts. The possible answers in terms of variations for a single phenotype are not limited and every user can add completely new phenotypes if the corresponding questions about this are lacking.

On the side of getting access to the data users can download single genotyping files for specific users, get archives of multiple genotyping files grouped by phenotypic variation or can access a single download that includes all genotyping-files and all phenotypes in a comma separated table. Additionally users can access the genetic data through the Distributed Annotation System, which allows to get all data for specific chromosomes and specific positions on single chromosomes.

Between the start of openSNP on 09/29/2011 and 12/17/2011 214 people have signed up with openSNP, 79 of those have uploaded their genotyping results. In the same timeframe all users combined have entered 675 variations which are distributed over 47 different phenotypes. The phenotype which has been provided by the most users is the eye color which has been provided by 54 different users. There are two phenotypes which have only been provided by a single user: The score of the Writing-SAT and triglyceride-levels.

\section*{Discussion}

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Materials and Methods}
\subsection*{Survey on Sharing Genetic Information}
The survey was done with Google Docs and included questions on the age, chromosomal sex and ethnicity of the participants. Additionally it included questions on if they are already customer of a DTC company, are planning to become one or don't plan to become one. If they are already a customer they also got asked if they already share their genetic and phenotypic data. All participants got asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, but only if they did not share my medical information with anybody else" and No".

The survey also asked some scale-questions which measured how strong participants agree/disagree on different reasons to share or not to share their information with third parties. The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree. Motivations given to share data were "because you want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease you have, possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations given not to share data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", "because of the breach of your privacy" and " because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". Additionally participants had the possibility to give own reasons to share or not to share their data.

The data of the survey was analyzed with SPSS 19.

\subsection*{Technical realization of the platform}
The main platform is realized using the web-framework Ruby on Rails. Postgres is used as the main database backend for Rails. The database stores genotyping results, phenotypic information of the users, literature results of Mendeley, the Public Library of Science and summaries on SNPs which can be found in SNPedia. The literature database of Mendeley is queried using the REST API, which delivers results in JSON. The literature database of the Public Library of Science is queried using the respective REST API, which delivers results in an XML-format. Summaries on SNPs are provided by SNPedia, through querying the content through the MediaWiki-API.

SNPs are cataloged by their unique identifier, which consists of a prefix (mostly \textit{rs}) and a unique number. This is a common format which is used by the NCBI dbSNP database and is also widely used and easily parsed from different literature-sources.

Processes with a longer runtime, such as parsing the genotyping results, creating archives of results which are be mailed to users and queries to external resources are handled using the ruby-gem Resque and a Redis-server. Search-features on the platform itself are implemented using SOLR and the ruby-gem sunspot. Additionally data can be requested from openSNP using the Distributed Annotation System. The data for this is stored in a mySQL-database, the delivery of the data is done by using ProServer, which was modified by Gel et al. for use in easyDAS.

The source code of openSNP is published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
